---
title: "Optimal Location of Low-Income Housing"
author: "Nino Martinez"
subtitle: "Gentrification, Displacement, and Accessibility"
---

# Introduction  

The purpose of this project is to use census data and recently available public data to  assist in the location of project based housing in Buffalo, NY’s disinvested and gentrifying neighborhoods. Through a combination of Geographic Information Systems (GIS), Operational Research, and Data Analytics, we legitimize the potential use of residential vacant land with Location Science, while considering equitable public transportation access.

Research Questions :

Given the rising market value of homes in some neighborhoods in Buffalo and potential for displacement:

1. How can city planners mitigate the extent of displacement?

2. How can vacant residential parcels be used to build new low-income housing?

3. Which land parcels will optimize the location of affordable housing to minimize the total weighted distance to the nearest metro-rail station for equitable accessibility to Public Transportation?


  
# Data 
1. To Analyse Concentrated Disadvantage for preliminary analysis of gentrifiable census tracts
- **Source:** [Neighborhood Change Database (NCDB)](http://www.geolytics.com/USCensus,Neighborhood-Change-Database-1970-2000,Products.asp)

-*Decline70_00.csv*

-**Source:** [U.S. Census Tiger Data](https://www.census.gov/geo/maps-data/data/tiger.html)

- *Buffalo.shp*
        
-**Source:** [Open data Buffalo Portal](https://data.buffalony.gov/)

- *neighborhoods.shp*

2. To Quantify Gentrification
  - 2018-2019 Assessment Roll (.csv)
  - Neighborhood boundaries (.shp)
3. Census Data
  - 2010 U.S. Census Bureau American Community Survey 5-year Estimates (2006-2010)
  - 2016 U.S. Census Bureau American Community Survey 5-year Estimates (2012-2016)
  - City census tract boundaries (.shp)
  
# Methodology

1. Identify disinvested neighborhoods based on concentrated disadvantage and shrinkage

2. Identify gentrifying neighborhoods based on assessment data

3. Identify Displacement Risk

4. P-median problem for vacant residential parcels and rail stations

- Euclidean Distance
- Public Transportation Usage
- Displacement Risk based off of 2016 census data

## Measuring Concentrated Disadvantage

In order to properly quantify gentrification, it is necessary to measure urban decline and concentrated disadvantage within the city historically. “By definition, gentrification is reinvestment in and rebuilding of the physical structures that have undergone a period of disinvestment" (Heidkamp & Lucas, 107). Neighborhoods that experience a period of disinvesment are often concentrated in disadvantage. Circumstances that make success (most often economically) unusually difficult make places disadvantaged, and “the presence of disadvantage weakens a place’s ability to achieve this goal [economic success]” (Weaver et al, 35). When multiple layers of disadvantage occur in a single location, concentrated disadvantage exists (36). By using the geometric mean of five disadvantage variables provided by the NCDB, I calculate concentrate disadvantage (CD) for the years 1970-2000. I use the same variables used by Weaver et al, 2017 in my calculations because all variables are represented as percentages, thus making comparisons across time consistent as they range from a value of 0-100. The equation is represented as:

**$Gi  = [(Percent Nonwhitei)* (Percent Female Headed Householdsi) * (Percent Unemployedi) * (Percent in Povertyi) * (Percent Low Educationi)]^1/5, i: 1, 2, …, n$**

```{r message = FALSE, warning=FALSE}
#load all necessary libraries for analysis
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tmap)
library(broom)
library(ggmap)
library(sp)
library(sf) 
library(reshape2)
library(rgeos)
library(rgdal)
library(tbart)
require(RColorBrewer)
require(GISTools)
library(leaflet)

```

## Measuring Gentrification

```{r}

# Load all data previously calculated

load("buff_CD_shrink.Rdata")
load("buffalo_2010.Rdata")
load("buffalo_2016.Rdata")
load("buffalo_change.Rdata")
load("gent_frontier.Rdata")
load("disinvested_gent.Rdata")
load("displace_sold_df.Rdata")
load("property_summary.Rdata")
neighborhoods <- st_read("Data/nbhds.shp")
```

## Measuring Displacement

```{r}

assess <- read.csv("assessment18_19.csv")

# Filter tax parcels for vacant residential that are geocoded


vacant_analysis <- c("RESIDENTIAL VACANT LAND", "RESIDENTIAL VACANT LAND WITH SMALL IMPROVEMENTS")

vacant_assess <- assess %>% filter(PROPERTY.CLASS.DESCRIPTION %in% vacant_analysis) 


vacant_geocode <- vacant_assess[!(is.na(vacant_assess$CENSUS.TRACT) | vacant_assess$CENSUS.TRACT==""),]


vacant_buff <- vacant_geocode %>% filter(OWNER1 == "CITY OF BUFFALO")


# Convert vacant dataframe into spatial object

buff_2016 <- as.data.frame(buffalo_2016_df)

# Displacement Risk

disinvested_rank <- disinvested_gent[,c(4,25,40:61)] 
disinvested_rank_df <- as.data.frame(disinvested_rank)
gent_displace <- full_join(displace_sold_df, disinvested_rank_df, by = c("GEOID10"= "GEOID10")) %>%
  mutate(diff_sale_value = sale_after - value_before,
         per_sale_value = (diff_sale_value/value_before)*100)

gent_summary <- full_join(gent_displace, property_summary, by = c("NAME10.x"="CENSUS.TRACT")) %>% 
  filter(!is.na(NAME10.y))

PT <- buff_2016[,c(4,13)]

gent_summary.2 <- full_join(gent_displace, PT, by = c("GEOID10"= "GEOID10"))


gent_summary_rank <- gent_summary.2 %>% mutate(gent_white = white_change*157,
                                             gent_pov = Per_Pov_Chg*165,
                                             gent_adult = per_25_44_change*239,
                                             gent_college = college_change*-429,
                                             gent_own = own_occ_change_per*-526,
                                             gent_occ = Per_Occ_Mgt_Chg*-526,
                                             gent_pop = total_pop_chg*-2.72,
                                             gent_rent = rent_change*51.6,
                                             value_added = gent_white +
                                               gent_pov + gent_adult + gent_college +
                                               gent_own + gent_pop + gent_rent) %>%
  filter(value_added > 0)
                                  
value_added_map <- tm_shape(gent_summary_rank) + tm_polygons("value_added", style = "jenks",
                                         palette = "seq",
                                         n = 7,
                                         title = "Value Added",
                                         border.alpha = .50) + 
  tm_shape(neighborhoods) + tm_polygons(alpha = 0, border.col = "black") +
  tm_layout(aes.palette = list(seq ="YlGn"), legend.outside = TRUE, main.title = "Value Added", main.title.size = .75)
value_added_map
                                          
                                                 

risk <- gent_summary_rank %>% mutate(own_high =
                                   ifelse(SMOCAPI_35plus_per_16 < 15.88, 1,
                                          ifelse(between(SMOCAPI_35plus_per_16, 15.88, 21.65), 2,
                                                 ifelse(between(SMOCAPI_35plus_per_16,21.65,31.27), 3,
                                                        ifelse(SMOCAPI_35plus_per_16 > 31.27,4,0)))),
                                 rent_high =
                                   ifelse(GRAPI_35plus_per_16 < 38.30, 1,
                                          ifelse(between(GRAPI_35plus_per_16,38.30,48.60), 2,
                                                 ifelse(between(GRAPI_35plus_per_16,48.60,58.60), 3,
                                                        ifelse(GRAPI_35plus_per_16 > 58.60, 4, 0)))),
                                 long_own =
                                   ifelse(move_in_27plus_yrs_per_16 < 11.03, 1,
                                          ifelse(between(move_in_27plus_yrs_per_16, 11.03,14.35),2,
                                                 ifelse(between(move_in_27plus_yrs_per_16,14.35,22.82),3,
                                                        ifelse(move_in_27plus_yrs_per_16 > 22.82, 4,0)))),
                                 renters =
                                   ifelse(rent_occ_per_16 < 50.70, 1,
                                          ifelse(between(rent_occ_per_16,50.7,59.8), 2,
                                                 ifelse(between(rent_occ_per_16,59.8,65.85),3,
                                                        ifelse(rent_occ_per_16 > 65.85,4,0)))),
                                 elderly =
                                   ifelse(per_65_plus_16 < 8.113, 1,
                                          ifelse(between(per_65_plus_16,8.113,12.229), 2,
                                                 ifelse(between(per_65_plus_16,12.229,15.165), 3,
                                                        ifelse(per_65_plus_16 > 15.165, 4, 0)))),
                                 market_value =
                                   ifelse(per_sale_change < 1.937, 1,
                                          ifelse(between(per_sale_change, 1.937,15.590), 2,
                                                 ifelse(between(per_sale_change,15.590,62.186),3,
                                                        ifelse(per_sale_change > 62.186, 4, 0)))),
                                 disp.rank =  renters + elderly + long_own + rent_high + own_high,
                                 risk = diff_sale_value * disp.rank) %>%
  filter(risk > 0)





value_prior_map <- tm_shape(gent_summary_rank) + tm_polygons("value_before", style = "jenks",
                                         palette = "seq",
                                         n = 7,
                                         title = "Avg Home Value",
                                         border.alpha = .50) + 
  tm_shape(neighborhoods) + tm_polygons(alpha = 0, border.col = "black") +
  tm_layout(aes.palette = list(seq ="YlGn"), legend.outside = TRUE, main.title = "Sold Prior 2011", main.title.size = .75)

value_prior_map

sold_after_map <- tm_shape(gent_summary_rank) + tm_polygons("sale_after", style = "jenks",
                                                             palette = "seq",
                                                             n = 7,
                                                             title = "Avg Sale Price",
                                                             border.alpha = .50) + 
  tm_shape(neighborhoods) + tm_polygons(alpha = 0, border.col = "black") +
  tm_layout(aes.palette = list(seq ="YlGn"), legend.outside = TRUE, main.title = "Sold After 2011", main.title.size = .75)

sold_after_map

per_chg_map <- tm_shape(gent_summary_rank) + tm_polygons("per_sale_value", style = "jenks",
                                                            palette = "div",
                                                            n = 7,
                                                            title = "% Change",
                                                            border.alpha = .50) + 
  tm_shape(neighborhoods) + tm_polygons(alpha = 0, border.col = "black") +
  tm_layout(aes.palette = list(seq = "plasma", div = c("BrBG")), legend.outside = TRUE, main.title = "Value Before and Sold After Assessment", main.title.size = .75)

per_chg_map



disp_rank_map <- tm_shape(risk) + tm_polygons("disp.rank", style = "jenks",
                                         palette = "seq",
                                         n = 7,
                                         title = "Risk Rank",
                                         border.alpha = .50) + 
  tm_shape(neighborhoods) + tm_polygons(alpha = 0, border.col = "black") +
  tm_layout(aes.palette = list(seq ="OrRd"), legend.outside = TRUE, main.title = "Displacement Rank", main.title.size = .75)

disp_rank_map

risk_map <- tm_shape(risk) + tm_polygons("risk", style = "jenks",
                                         palette = "seq",
                                         n = 7,
                                         title = "% Change * Risk Rank",
                                         border.alpha = .50) + 
  tm_shape(neighborhoods) + tm_polygons(alpha = 0, border.col = "black") +
  tm_layout(aes.palette = list(seq ="OrRd"), legend.outside = TRUE, main.title = "Displacement Risk", main.title.size = .75)
risk_map
risk_sp <- as(risk, "Spatial")
# Join Risk dataframe with geocoded vacant df

risk_vacant <- full_join(vacant_geocode, risk, by = c("CENSUS.TRACT"="NAME10.x")) %>% 
  filter(!is.na(NAME10.y))

risk_vacant_buff <- risk_vacant %>% filter(OWNER1 == "CITY OF BUFFALO")

risk_vacant_coords <- risk_vacant[,c(45,44)]
risk_vacant_buff_coords <- risk_vacant_buff[,c(45,44)]

risk_vacant_sp <- SpatialPointsDataFrame(coords = risk_vacant_coords, data = risk_vacant,
                                         proj4string = CRS("+proj=tmerc +lat_0=40 +lon_0=-78.58333333333333 +k=0.9999375 +x_0=350000 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs") )

risk_vacant_buff_sp <- SpatialPointsDataFrame(coords = risk_vacant_buff_coords, data = risk_vacant_buff,
                                         proj4string = CRS("+proj=tmerc +lat_0=40 +lon_0=-78.58333333333333 +k=0.9999375 +x_0=350000 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs") )


#Try allocations model
tm_shape(risk_vacant_buff_sp) +tm_dots()
vacant__buff_allocation <- allocations(risk_vacant_buff_sp, p = 5)
vacant_buff_id <- unique(vacant__buff_allocation$allocation)



# allocations model with rail stops

rail <- st_read("Rail_Stations.shp")

rail_coords <- do.call(rbind, st_geometry(rail)) %>%
  as_tibble() %>% setNames(c("lon","lat"))

rail_sp <- SpatialPoints(coords = rail_coords,
                                  proj4string = CRS("+proj=tmerc +lat_0=40 +lon_0=-78.58333333333333 +k=0.9999375 +x_0=350000 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs") )

```

## Allocations Model for Disinvested Tracts

```{r}
## No weighted distance


allocation.a.1 <- allocations(risk_vacant_buff_sp, rail_sp, p = 1)
id.a.1 <- unique(allocation.a.1$allocation)
sites.a.1 <- risk_vacant_buff_sp[id.a.1,]
total.a.1 <- sum(allocation.a.1$allocdist)



allocation.a.2 <- allocations(risk_vacant_buff_sp, rail_sp, p = 2)
id.a.2 <- unique(allocation.a.2$allocation)
sites.a.2 <- risk_vacant_buff_sp[id.a.2,]
total.a.2 <- sum(allocation.a.2$allocdist)



allocation.a.3 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 3)
id.a.3 <- unique(allocation.a.3$allocation)
sites.a.3 <- risk_vacant_buff_sp[id.a.3,]
total.a.3 <- sum(allocation.a.3$allocdist)

allocation.a.4 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 4)
id.a.4 <- unique(allocation.a.4$allocation)
sites.a.4 <- risk_vacant_buff_sp[id.a.4,]
total.a.4 <- sum(allocation.a.4$allocdist)

allocation.a.5 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 5)
id.a.5 <- unique(allocation.a.5$allocation)
sites.a.5 <- risk_vacant_buff_sp[id.a.5,]
total.a.5 <- sum(allocation.a.5$allocdist)

allocation.a.6 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 6)
id.a.6 <- unique(allocation.a.6$allocation)
sites.a.6 <- risk_vacant_buff_sp[id.a.6,]
total.a.6 <- sum(allocation.a.6$allocdist)

allocation.a.7 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 7)
id.a.7 <- unique(allocation.a.7$allocation)
sites.a.7 <- risk_vacant_buff_sp[id.a.7,]
total.a.7 <- sum(allocation.a.7$allocdist)

allocation.a.8 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 8)
id.a.8 <- unique(allocation.a.8$allocation)
sites.a.8 <- risk_vacant_buff_sp[id.a.8,]
total.a.8 <- sum(allocation.a.8$allocdist)

allocation.a.9 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 9)
id.a.9 <- unique(allocation.a.9$allocation)
sites.a.9 <- risk_vacant_buff_sp[id.a.9,]
total.a.9 <- sum(allocation.a.9$allocdist)

allocation.a.10 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 10)
id.a.10 <- unique(allocation.a.10$allocation)
sites.a.10 <- risk_vacant_buff_sp[id.a.10,]
total.a.10 <- sum(allocation.a.10$allocdist)

allocation.a.11 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 11)
id.a.11 <- unique(allocation.a.11$allocation)
sites.a.11 <- risk_vacant_buff_sp[id.a.11,]
total.a.11 <- sum(allocation.a.11$allocdist)

allocation.a.12 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 12)
id.a.12 <- unique(allocation.a.12$allocation)
sites.a.12 <- risk_vacant_buff_sp[id.a.12,]
total.a.12 <- sum(allocation.a.12$allocdist)

allocation.a.13 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 13)
id.a.13 <- unique(allocation.a.13$allocation)
sites.a.13 <- risk_vacant_buff_sp[id.a.13,]
total.a.13 <- sum(allocation.a.13$allocdist)

allocation.a.14 <- allocations(risk_vacant_buff_sp, rail_sp,  p = 14)
id.a.14 <- unique(allocation.a.14$allocation)
sites.a.14 <- risk_vacant_buff_sp[id.a.14,]
total.a.14 <- sum(allocation.a.14$allocdist)

tradeoff_matrix.a <- matrix(nrow = 14, c(1:14, total.a.1, total.a.2, total.a.3, total.a.4, total.a.5, total.a.6, total.a.7,
                                         total.a.8, total.a.9, total.a.10, total.a.11, total.a.12, total.a.13, total.a.14))

tradeoff_df.a <- as.data.frame(tradeoff_matrix.a)
colnames(tradeoff_df.a) <- c("p", "Euclidean_Distance")

tradeoff_plot.a <- ggplot(tradeoff_df.a, mapping = aes(p, Euclidean_Distance)) +
  geom_line() +
  scale_x_continuous(breaks = c(1:15)) +
  labs(x = "Number of Facilities", y = "Euclidean Distance", title = "Tradeoff Curve")

tradeoff_plot.a


### Weighted by percentage take public transit

PT_matrix <- as.matrix(risk_vacant_buff_sp$Per_Public_Trans_16 *euc.dists(risk_vacant_buff_sp))

PT.1 <- allocations(risk_vacant_buff_sp, rail_sp, p = 1, metric = PT_matrix)
PT_id.1 <- unique(PT.1$allocation)
sites.c.1 <- PT.1[PT_id.1,]
total.c.1 <- sum(PT.1$allocdist)

PT.2 <- allocations(risk_vacant_buff_sp, rail_sp, p = 2, metric = PT_matrix)
PT_id.2 <- unique(PT.2$allocation)
sites.c.2 <- PT.1[PT_id.2,]
total.c.2 <- sum(PT.2$allocdist)

PT.3 <- allocations(risk_vacant_buff_sp, rail_sp, p = 3, metric = PT_matrix)
PT_id.3 <- unique(PT.3$allocation)
sites.c.3 <- PT.1[PT_id.3,]
total.c.3 <- sum(PT.3$allocdist)

PT.4 <- allocations(risk_vacant_buff_sp, rail_sp, p = 4, metric = PT_matrix)
PT_id.4 <- unique(PT.4$allocation)
sites.c.4 <- PT.4[PT_id.4,]
total.c.4 <- sum(PT.4$allocdist)

PT.5 <- allocations(risk_vacant_buff_sp, rail_sp, p = 5, metric = PT_matrix)
PT_id.5 <- unique(PT.5$allocation)
sites.c.5 <- PT.5[PT_id.5,]
total.c.5 <- sum(PT.5$allocdist)

PT.6 <- allocations(risk_vacant_buff_sp, rail_sp, p = 6, metric = PT_matrix)
PT_id.6 <- unique(PT.6$allocation)
sites.c.6 <- PT.6[PT_id.6,]
total.c.6 <- sum(PT.6$allocdist)

PT.7 <- allocations(risk_vacant_buff_sp, rail_sp, p = 7, metric = PT_matrix)
PT_id.7 <- unique(PT.7$allocation)
sites.c.7 <- PT.1[PT_id.7,]
total.c.7 <- sum(PT.7$allocdist)

PT.8 <- allocations(risk_vacant_buff_sp, rail_sp, p = 8, metric = PT_matrix)
PT_id.8 <- unique(PT.8$allocation)
sites.c.8 <- PT.8[PT_id.8,]
total.c.8 <- sum(PT.8$allocdist)

PT.9 <- allocations(risk_vacant_buff_sp, rail_sp, p = 9, metric = PT_matrix)
PT_id.9 <- unique(PT.9$allocation)
sites.c.9 <- PT.9[PT_id.9,]
total.c.9 <- sum(PT.9$allocdist)

PT.10 <- allocations(risk_vacant_buff_sp, rail_sp, p = 10, metric = PT_matrix)
PT_id.10 <- unique(PT.10$allocation)
sites.c.10 <- PT.10[PT_id.10,]
total.c.10 <- sum(PT.10$allocdist)

PT.11 <- allocations(risk_vacant_buff_sp, rail_sp, p = 11, metric = PT_matrix)
PT_id.11 <- unique(PT.11$allocation)
sites.c.11 <- PT.11[PT_id.11,]
total.c.11 <- sum(PT.11$allocdist)

PT.12 <- allocations(risk_vacant_buff_sp, rail_sp, p = 12, metric = PT_matrix)
PT_id.12 <- unique(PT.12$allocation)
sites.c.12 <- PT.12[PT_id.12,]
total.c.12 <- sum(PT.12$allocdist)

PT.13 <- allocations(risk_vacant_buff_sp, rail_sp, p = 13, metric = PT_matrix)
PT_id.13 <- unique(PT.13$allocation)
sites.c.13 <- PT.13[PT_id.13,]
total.c.13 <- sum(PT.13$allocdist)

PT.14 <- allocations(risk_vacant_buff_sp, rail_sp, p = 14, metric = PT_matrix)
PT_id.14 <- unique(PT.14$allocation)
sites.c.14 <- PT.14[PT_id.14,]
total.c.14 <- sum(PT.14$allocdist)

# Tradeoff Curve for Public Transportation

tradeoff_matrix.c <- matrix(nrow = 14, c(1:14, total.c.1, total.c.2, total.c.3, total.c.4, total.c.5, total.c.6, total.c.7,
                                         total.c.8, total.c.9, total.c.10, total.c.11, total.c.12, total.c.13, total.c.14))

tradeoff_df.c <- as.data.frame(tradeoff_matrix.c)
colnames(tradeoff_df.c) <- c("p", "Public_Trans")

tradeoff_plot.c <- ggplot(tradeoff_df.c, mapping = aes(p, Public_Trans)) +
  geom_line() +
  scale_x_continuous(breaks = c(1:14)) +
  labs(x = "Number of Facilities", y = "Distance Weighted Public Transit (%)", title = "Tradeoff Curve")

tradeoff_plot.c

## weighted by value added from regression

VA_matrix <- as.matrix(risk_vacant_buff_sp$value_added *euc.dists(risk_vacant_buff_sp))

VA.1 <- allocations(risk_vacant_buff_sp, rail_sp, p = 1, metric = VA_matrix)
VA_id.1 <- unique(VA.1$allocation)
sites.d.1 <- VA.1[VA_id.1,]
total.d.1 <- sum(VA.1$allocdist)

VA.2 <- allocations(risk_vacant_buff_sp, rail_sp, p = 2, metric = VA_matrix)
VA_id.2 <- unique(VA.2$allocation)
sites.d.2 <- VA.2[VA_id.2,]
total.d.2 <- sum(VA.2$allocdist)

VA.3 <- allocations(risk_vacant_buff_sp, rail_sp, p = 3, metric = VA_matrix)
VA_id.3 <- unique(VA.3$allocation)
sites.d.3 <- VA.3[VA_id.3,]
total.d.3 <- sum(VA.3$allocdist)

VA.4 <- allocations(risk_vacant_buff_sp, rail_sp, p = 4, metric = VA_matrix)
VA_id.4 <- unique(VA.4$allocation)
sites.d.4 <- VA.4[VA_id.4,]
total.d.4 <- sum(VA.4$allocdist)

VA.4 <- allocations(risk_vacant_buff_sp, rail_sp, p = 4, metric = VA_matrix)
VA_id.4 <- unique(VA.4$allocation)
sites.d.4 <- VA.4[VA_id.4,]
total.d.4 <- sum(VA.4$allocdist)

VA.5 <- allocations(risk_vacant_buff_sp, rail_sp, p = 5, metric = VA_matrix)
VA_id.5 <- unique(VA.5$allocation)
sites.d.5 <- VA.5[VA_id.5,]
total.d.5 <- sum(VA.5$allocdist)

VA.6 <- allocations(risk_vacant_buff_sp, rail_sp, p = 6, metric = VA_matrix)
VA_id.6 <- unique(VA.6$allocation)
sites.d.6 <- VA.6[VA_id.6,]
total.d.6 <- sum(VA.6$allocdist)

VA.7 <- allocations(risk_vacant_buff_sp, rail_sp, p = 7, metric = VA_matrix)
VA_id.7 <- unique(VA.7$allocation)
sites.d.7 <- VA.7[VA_id.7,]
total.d.7 <- sum(VA.7$allocdist)

VA.8 <- allocations(risk_vacant_buff_sp, rail_sp, p = 8, metric = VA_matrix)
VA_id.8 <- unique(VA.8$allocation)
sites.d.8 <- VA.8[VA_id.8,]
total.d.8 <- sum(VA.8$allocdist)

VA.9 <- allocations(risk_vacant_buff_sp, rail_sp, p = 9, metric = VA_matrix)
VA_id.9 <- unique(VA.9$allocation)
sites.d.9 <- VA.9[VA_id.9,]
total.d.9 <- sum(VA.9$allocdist)

VA.10 <- allocations(risk_vacant_buff_sp, rail_sp, p = 10, metric = VA_matrix)
VA_id.10 <- unique(VA.10$allocation)
sites.d.10 <- VA.10[VA_id.10,]
total.d.10 <- sum(VA.10$allocdist)

VA.11 <- allocations(risk_vacant_buff_sp, rail_sp, p = 11, metric = VA_matrix)
VA_id.11 <- unique(VA.11$allocation)
sites.d.11 <- VA.11[VA_id.11,]
total.d.11 <- sum(VA.11$allocdist)

VA.12 <- allocations(risk_vacant_buff_sp, rail_sp, p = 12, metric = VA_matrix)
VA_id.12 <- unique(VA.12$allocation)
sites.d.12 <- VA.12[VA_id.12,]
total.d.12 <- sum(VA.12$allocdist)

VA.13 <- allocations(risk_vacant_buff_sp, rail_sp, p = 13, metric = VA_matrix)
VA_id.13 <- unique(VA.13$allocation)
sites.d.13 <- PT.13[VA_id.13,]
total.d.13 <- sum(VA.13$allocdist)

VA.14 <- allocations(risk_vacant_buff_sp, rail_sp, p = 14, metric = VA_matrix)
VA_id.14 <- unique(VA.14$allocation)
sites.d.14 <- VA.14[VA_id.14,]
total.d.14 <- sum(VA.14$allocdist)

# Tradeoff curve weighted by value added

tradeoff_matrix.d <- matrix(nrow = 14, c(1:14, total.d.1, total.d.2, total.d.3, total.d.4, total.d.5, total.d.6, total.d.7,
                                         total.d.8, total.d.9, total.d.10, total.d.11, total.d.12, total.d.13, total.d.14))

tradeoff_df.d <- as.data.frame(tradeoff_matrix.d)
colnames(tradeoff_df.d) <- c("p", "value_added")

tradeoff_plot.d <- ggplot(tradeoff_df.d, mapping = aes(p, value_added)) +
  geom_line() +
  scale_x_continuous(breaks = c(1:14)) +
  labs(x = "Number of Facilities", y = "Distance Weighted Value Added ($)", title = "Tradeoff Curve")

tradeoff_plot.d

## weighted by displacement risk

DR_matrix <- as.matrix(risk_vacant_buff_sp$risk *euc.dists(risk_vacant_buff_sp))

DR.1 <- allocations(risk_vacant_buff_sp, rail_sp, p = 1, metric = DR_matrix)
DR_id.1 <- unique(DR.1$allocation)
sites.e.1 <- DR.1[DR_id.1,]
total.e.1 <- sum(DR.1$allocdist)

DR.2 <- allocations(risk_vacant_buff_sp, rail_sp, p = 2, metric = DR_matrix)
DR_id.2 <- unique(DR.2$allocation)
sites.e.2 <- DR.2[DR_id.2,]
total.e.2 <- sum(DR.2$allocdist)

DR.3 <- allocations(risk_vacant_buff_sp, rail_sp, p = 3, metric = DR_matrix)
DR_id.3 <- unique(DR.3$allocation)
sites.e.3 <- DR.3[DR_id.3,]
total.e.3 <- sum(DR.3$allocdist)

DR.4 <- allocations(risk_vacant_buff_sp, rail_sp, p = 4, metric = DR_matrix)
DR_id.4 <- unique(DR.4$allocation)
sites.e.4 <- DR.4[DR_id.4,]
total.e.4 <- sum(DR.4$allocdist)

DR.5 <- allocations(risk_vacant_buff_sp, rail_sp, p = 5, metric = DR_matrix)
DR_id.5 <- unique(DR.5$allocation)
sites.e.5 <- DR.5[DR_id.5,]
total.e.5 <- sum(DR.5$allocdist)

DR.6 <- allocations(risk_vacant_buff_sp, rail_sp, p = 6, metric = DR_matrix)
DR_id.6 <- unique(DR.6$allocation)
sites.e.6 <- DR.6[DR_id.6,]
total.e.6 <- sum(DR.6$allocdist)

DR.7 <- allocations(risk_vacant_buff_sp, rail_sp, p = 7, metric = DR_matrix)
DR_id.7 <- unique(DR.7$allocation)
sites.e.7 <- DR.7[DR_id.7,]
total.e.7 <- sum(DR.7$allocdist)

DR.8 <- allocations(risk_vacant_buff_sp, rail_sp, p = 8, metric = DR_matrix)
DR_id.8 <- unique(DR.8$allocation)
sites.e.8 <- DR.8[DR_id.8,]
total.e.8 <- sum(DR.8$allocdist)

DR.9 <- allocations(risk_vacant_buff_sp, rail_sp, p = 9, metric = DR_matrix)
DR_id.9 <- unique(DR.9$allocation)
sites.e.9 <- DR.9[DR_id.9,]
total.e.9 <- sum(DR.9$allocdist)

DR.10 <- allocations(risk_vacant_buff_sp, rail_sp, p = 10, metric = DR_matrix)
DR_id.10 <- unique(DR.10$allocation)
sites.e.10 <- DR.10[DR_id.10,]
total.e.10 <- sum(DR.10$allocdist)

DR.11 <- allocations(risk_vacant_buff_sp, rail_sp, p = 11, metric = DR_matrix)
DR_id.11 <- unique(DR.11$allocation)
sites.e.11 <- DR.11[DR_id.11,]
total.e.11 <- sum(DR.11$allocdist)

DR.12 <- allocations(risk_vacant_buff_sp, rail_sp, p = 12, metric = DR_matrix)
DR_id.12 <- unique(DR.12$allocation)
sites.e.12 <- DR.12[DR_id.12,]
total.e.12 <- sum(DR.12$allocdist)

DR.13 <- allocations(risk_vacant_buff_sp, rail_sp, p = 13, metric = DR_matrix)
DR_id.13 <- unique(DR.13$allocation)
sites.e.13 <- DR.13[DR_id.13,]
total.e.13 <- sum(DR.13$allocdist)

DR.14 <- allocations(risk_vacant_buff_sp, rail_sp, p = 14, metric = DR_matrix)
DR_id.14 <- unique(DR.14$allocation)
sites.e.14 <- DR.14[DR_id.14,]
total.e.14 <- sum(DR.14$allocdist)

# Tradeoff curve for weighted by displacement risk

tradeoff_matrix.e <- matrix(nrow = 14, c(1:14, total.e.1, total.e.2, total.e.3, total.e.4, total.e.5, total.e.6, total.e.7,
                                         total.e.8, total.e.9, total.e.10, total.e.11, total.e.12, total.e.13, total.e.14))

tradeoff_df.e <- as.data.frame(tradeoff_matrix.e)
colnames(tradeoff_df.e) <- c("p", "DR")

tradeoff_plot.e <- ggplot(tradeoff_df.e, mapping = aes(p, DR)) +
  geom_line() +
  scale_x_continuous(breaks = c(1:14)) +
  labs(x = "Number of Facilities", y = "Distance Weighted Displacment Risk ($)", title = "Tradeoff Curve")

tradeoff_plot.e

## all optimal sites




## Map for 14 sites


DR_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  DR.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)

VA_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  VA.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)

PT_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  PT.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)

a_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  allocation.a.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)


p.14_map <- leaflet() %>% setView(-78.8784, 42.8864, 10.9) %>% addTiles() %>%
  addCircleMarkers(data = DR.14,
                   lng = DR.14$LONGITUDE,
                   lat = DR.14$LATITUDE,
                   radius = 2,
                   fillOpacity = .7,
                   color = "Black",
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "Potential Sites") %>%
  addCircleMarkers(data = sites.e.14,
                   lng = sites.e.14$LONGITUDE,
                   lat = sites.e.14$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "Displacement Risk") %>%
  addCircleMarkers(data = sites.d.14,
                   lng = sites.d.14$LONGITUDE,
                   lat = sites.d.14$LATITUDE,
                   radius = 4,
                   color = "#FF7F00",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = VA_p.14_labels,
                   group = "Value Added") %>%
  addCircleMarkers(data = sites.c.14,
                   lng = sites.c.14$LONGITUDE,
                   lat = sites.c.14$LATITUDE,
                   radius = 4,
                   color = "#00FF00",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = PT_p.14_labels,
                   group = "Public Transportation") %>%
  addCircleMarkers(data = sites.a.14,
                   lng = sites.a.14$LONGITUDE,
                   lat = sites.a.14$LATITUDE,
                   radius = 4,
                   color = "#0000FF",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = a_p.14_labels,
                   group = "Euclidean Distance") %>%
   addLayersControl(baseGroups = "Potential Sites",
     overlayGroups = c("Displacement Risk","Value Added", "Public Transportation", "Euclidean Distance"),
                   options = layersControlOptions(collapsed = FALSE))


p.14_map

p.7_map <- leaflet() %>% setView(-78.8784, 42.8864, 10.9) %>% addTiles() %>%
  addCircleMarkers(data = DR.7,
                   lng = DR.7$LONGITUDE,
                   lat = DR.7$LATITUDE,
                   radius = 2,
                   fillOpacity = .7,
                   color = "Black",
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "Potential Sites") %>%
  addCircleMarkers(data = sites.e.7,
                   lng = sites.e.7$LONGITUDE,
                   lat = sites.e.7$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "Displacement Risk") %>%
  addCircleMarkers(data = sites.d.7,
                   lng = sites.d.7$LONGITUDE,
                   lat = sites.d.7$LATITUDE,
                   radius = 4,
                   color = "#FF7F00",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = VA_p.14_labels,
                   group = "Value Added") %>%
  addCircleMarkers(data = sites.c.7,
                   lng = sites.c.7$LONGITUDE,
                   lat = sites.c.7$LATITUDE,
                   radius = 4,
                   color = "#00FF00",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = PT_p.14_labels,
                   group = "Public Transportation") %>%
  addCircleMarkers(data = sites.a.7,
                   lng = sites.a.7$LONGITUDE,
                   lat = sites.a.7$LATITUDE,
                   radius = 4,
                   color = "#0000FF",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = a_p.14_labels,
                   group = "Euclidean Distance") %>%
  addLayersControl(baseGroups = "Potential Sites",
                   overlayGroups = c("Displacement Risk","Value Added", "Public Transportation", "Euclidean Distance"),
                   options = layersControlOptions(collapsed = FALSE))


p.7_map

## Map for displacement risk


DR_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  DR.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)

VA_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  VA.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)

PT_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  PT.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)

a_p.14_labels <- sprintf(
  "<strong>%g Total Value</strong>",
  allocation.a.14$TOTAL.VALUE) %>% lapply(htmltools::HTML)


DR_all_map <- leaflet() %>% setView(-78.8784, 42.8864, 10.9) %>% addTiles() %>%
  addCircleMarkers(data = DR.7,
                   lng = DR.7$LONGITUDE,
                   lat = DR.7$LATITUDE,
                   radius = 2,
                   fillOpacity = .7,
                   color = "Black",
                   stroke = FALSE,
                   group = "Potential Sites") %>%
  addCircleMarkers(data = sites.e.1,
                   lng = sites.e.1$LONGITUDE,
                   lat = sites.e.1$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 1") %>%
  addCircleMarkers(data = sites.e.2,
                   lng = sites.e.2$LONGITUDE,
                   lat = sites.e.2$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 2") %>%
  addCircleMarkers(data = sites.e.3,
                   lng = sites.e.3$LONGITUDE,
                   lat = sites.e.3$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 3") %>%
  addCircleMarkers(data = sites.e.4,
                   lng = sites.e.4$LONGITUDE,
                   lat = sites.e.4$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 4") %>%
  addCircleMarkers(data = sites.e.5,
                   lng = sites.e.5$LONGITUDE,
                   lat = sites.e.5$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 5") %>%
  addCircleMarkers(data = sites.e.6,
                   lng = sites.e.6$LONGITUDE,
                   lat = sites.e.6$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 6") %>%
  addCircleMarkers(data = sites.e.7,
                   lng = sites.e.7$LONGITUDE,
                   lat = sites.e.7$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 7") %>%
  addCircleMarkers(data = sites.e.8,
                   lng = sites.e.8$LONGITUDE,
                   lat = sites.e.8$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 8") %>%
  addCircleMarkers(data = sites.e.9,
                   lng = sites.e.9$LONGITUDE,
                   lat = sites.e.9$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 9") %>%
  addCircleMarkers(data = sites.e.10,
                   lng = sites.e.10$LONGITUDE,
                   lat = sites.e.10$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 10") %>%
  addCircleMarkers(data = sites.e.11,
                   lng = sites.e.11$LONGITUDE,
                   lat = sites.e.11$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 11") %>%
  addCircleMarkers(data = sites.e.12,
                   lng = sites.e.12$LONGITUDE,
                   lat = sites.e.12$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 12") %>%
  addCircleMarkers(data = sites.e.13,
                   lng = sites.e.13$LONGITUDE,
                   lat = sites.e.13$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 13") %>%
  addCircleMarkers(data = sites.e.14,
                   lng = sites.e.14$LONGITUDE,
                   lat = sites.e.14$LATITUDE,
                   radius = 4,
                   color = "Red",
                   fillOpacity = .9,
                   stroke = FALSE,
                   label = DR_p.14_labels,
                   group = "p = 14") %>%
  addLayersControl(baseGroups = "Potential Sites",
                   overlayGroups = c("p = 1", "p = 2", "p = 3",
                                     "p = 4", "p = 5", "p = 6",
                                     "p = 7", "p = 8", "p = 9",
                                     "p = 10", "p = 11", "p = 12",
                                     "p = 13", "p = 14"),
                   options = layersControlOptions(collapsed = TRUE))


DR_all_map



```

# Conclusions

What have you learned?  Are there any broader implications?

# References 

Freeman, Lance (2009). Neighbourhood Diversity, Metropolitan Segregation and Gentrification: What Are the Links in the US? Urban Studies 46(10), 2079-2101.

Holm, Andrej & Schulz, Guido (2017). Gentrimap: A Model for Measuring Gentrification and Displacement. Gentrification and Resistance: Researching Displacement Processes and Adaption Strategies. Springer VS, 251-277

Heidkamp, Patrick C & Lucas, Susan (2006). Finding the Gentrification Frontier Using Census Data: The Case of Portland, Maine. Urban Geography 27(2), 101-125.

LeGates, R., & Hartman, C. (1982). Gentrification-Caused Displacement. The Urban Lawyer, 14(1), 31-55. 

Zook et al (2017). Big Data and the City. Handbook of Urban Geography, 1-12.

